A preliminary constraint grammar for Russian

### 2015-05-07 2015-nodalida-ruscg/presentation_ruscg.tex
more slides
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60293 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{frame}
\frametitle{New finite-state tools for Russian morphological analysis}
\framesubtitle{}
\begin{itemize}
	\item We developed a finite-state transducer using the two-level formalism \cite{Koskenniemi-84}
	\begin{itemize}	
		\item analyze and generate Russian wordforms, with or without stress marks
		\pause
	\end{itemize}
	\item Based on Zaliznjak's \emph{Grammatical dictionary of Russian} \cite{Zaliznjak-77}
	\begin{itemize}
		\item includes the proper noun appendix from 2001 version
		\item supplemented with \emph{Grammatical dictionary of new Russian words} [Grishina \& Lyashevskaya 2008]
		\pause
	\end{itemize}
	\item Since Russian has systematic syncretism and widespread homonymy, we are developing a constraint grammar \cite{Karlsson.Voutilainen.ea-95} to disambiguate such cases
	\pause
	\item implemented in the freely available CG3 system (http://beta.visl.sdu.dk/cg3.html)
\end{itemize}
\end{frame}

@@

\section{Development process} %FRAN
\begin{frame}
\frametitle{}
\framesubtitle{}
\begin{itemize}
	\item 
	\pause
	\item 
	\begin{itemize}
		\item 
		\pause
	\end{itemize}
	\item 
\end{itemize}
\end{frame}

>>>>

\begin{frame}
\frametitle{New finite-state transducer for Russian morphological analysis}
\framesubtitle{}
\begin{itemize}
	\item We developed a finite-state transducer using the two-level formalism \cite{Koskenniemi-84}
	\begin{itemize}
		\item lexc/twolc, can be compiled in xfst or hfst
		\item analyze and generate Russian wordforms, with or without stress marks
		\pause
	\end{itemize}
	\item Based on Zaliznjak's \emph{Grammatical dictionary of Russian} \cite{Zaliznjak-77}
	\begin{itemize}
		\item includes the proper noun appendix from 2001 version
		\item supplemented with \emph{Grammatical dictionary of new Russian words} [Grishina \& Lyashevskaya 2008]
		\pause
	\end{itemize}
	\item Example output:
	\begin{itemize}
		\item \rus{новый}\texttt{{\small <adj><m><nn><sg><nom>}}\\
	    `new'
		\item \rus{автомат}\texttt{{\small <n><m><nn><sg><nom>}}\\
	    `automaton, sub-machine gun'
	\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{CG rules: distribution of rule across levels}
\framesubtitle{}
\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
  \label{tab:ruleDist}
\end{table}
\begin{itemize}
	\item Safe: real constraints in the language\\
		{\small Ex: preposition cannot directly precede a finite verb}
		\pause
	\item Safe heuristic: highly reliable tendencies\\ % Russian word order is very free
		{\small Ex: remove GEN from sentence-init. if no verb governing GEN.}
		\pause
	\item Heuristic: tendencies that are true most of the time\\ % better replaced with machine-learning
		{\small Ex: \rus{такая} is `such.FEM-SG-NOM', not `say~\rus{так}.V-PRS-ACT-ADV'}
\end{itemize}
\end{frame}

\section{Development process} %FRAN

\begin{frame}
\frametitle{Writing rules and level assignment}
\framesubtitle{}
\begin{itemize}
	\item No hand-annotated corpus was compatible with our analyser
	\pause
	\item We sorted new rules using Russian Wikipedia by collecting 100 example applications of the rule
	\pause
	\begin{itemize}
		\item Safe: 100\% correct
		\item Safe heuristic: > 97\% correct
		\item Heuristic (or discard): $\leq$ 97\% correct 
	\end{itemize}
\end{itemize}
\end{frame}
